Supporting 10 common mistakes article

### 2025-08-29 example_10m.Rmd
using data to support the 10 mistakes article

@@

## Enrichment analysis with functional class scoring

| Reporting criteria | Method/resource used |
| --- | --- |
| Origin of gene sets | Reactome (2023-03-06) |
| Tool used | FGSEA (check version at foot of report) |
| Statistical test used | Kolmogorov-Smirnov test |
| P-value correction for multiple comparisons | FDR method |
| Background list | Genes with >=10 reads per sample on average across all samples |
| Gene Selection Criteria | DESeq2 FDR<0.05 |
| ORA directionality | Separate tests for up- and down-regulation |
| Data availability | via DEE2 at accession SRP038101 (human) |
| Other parameters | Min gene set size of 10 |

```{r,fgsea}

@@

```{r,topbarplot3}

par(mar=c(5,20,5,1))

cols <- c(rep("blue",n_pw),rep("red",n_pw))

barplot(c(fgsea_dn_vec,fgsea_up_vec),
  horiz=TRUE,las=1,cex.names=0.65,col=cols,
  main="top DE Reactomes",
  xlab="ES",
  cex.main=0.9)

mtext("FCS test")

>>>>

suppressPackageStartupMessages({
  library("getDEE2")
  library("DESeq2")
  library("kableExtra")
  library("clusterProfiler")
  library("eulerr")
  library("gplots")
  library("mitch")
  library("fgsea")
  library("RhpcBLASctl")
})

@@

## Mistake 1 - Raw p-values

## Mistake 2 - Background

## Mistake 4 - prioritisation by p-values

| Reporting criteria | Method/resource used |
| --- | --- |
| Origin of gene sets | Reactome (2023-03-06) |
| Tool used | FORA (check fgsea version at foot of report) |
| Statistical test used | hypergeometric |
| P-value correction for multiple comparisons | FDR method |
| Background list | Genes with >=10 reads per sample on average across all samples |
| Gene Selection | DESeq2 FDR<0.05 and log2FC>0 or log2FC<0 |
| Data availability | via DEE2 at accession SRP038101 (human) |
| Other parameters | Min gene set size of 5 |

```{r,mistake2}

up <- unique(sapply(strsplit(rownames(subset(def, log2FoldChange>0 & padj<0.05))," "),"[[",2))

dn <- unique(sapply(strsplit(rownames(subset(def, log2FoldChange<0 & padj<0.05))," "),"[[",2))

bg <- unique(sapply(strsplit(rownames(def)," "),"[[",2))

gt <- read.table("https://dee2.io/data/hsapiens/hsa_gene_info.tsv",header=TRUE)
head(gt)

all <- unique(gt$GeneSymbol)

lapply(list(up,dn,bg,all),length)

fres_up_bg <- fora(pathways=gsets, genes=up, universe=bg, minSize = 5, maxSize = Inf)
fres_up_bg <- subset(fres_up_bg,padj<0.05)$pathway
length(fres_up_bg)

fres_up_all <- fora(pathways=gsets, genes=up, universe=all, minSize = 5, maxSize = Inf)
fres_up_all <- subset(fres_up_all,padj<0.05)$pathway
length(fres_up_all)

fres_dn_bg <- fora(pathways=gsets, genes=dn, universe=bg, minSize = 5, maxSize = Inf)
fres_dn_bg <- subset(fres_dn_bg,padj<0.05)$pathway
length(fres_dn_bg)

fres_dn_all <- fora(pathways=gsets, genes=dn, universe=all, minSize = 5, maxSize = Inf)
fres_dn_all <- subset(fres_dn_all,padj<0.05)$pathway
length(fres_dn_all)

v1 <- list("up"=fres_up_bg, "up*"=fres_up_all, "dn"=fres_dn_bg, "dn*"=fres_dn_all)

par(mar=c(1,1,1,1))
par(mfrow=c(1,1))
plot(euler(v1),quantities = list(cex = 2), labels = list(cex = 2))

v2 <- list("up"=fres_up_bg, "up*"=fres_up_all)
plot(euler(v2),quantities = list(cex = 2), labels = list(cex = 2))

par(mar=c(5.1,4.1,4.1,2.1))

## Mistake 4 - prioritisation by p-values

| Reporting criteria | Method/resource used |
| --- | --- |
| Origin of gene sets | Reactome (2023-03-06) |
| Tool used | mitch (check version at foot of report) |
| Statistical test used | Rank ANOVA (Kruskal Wallis) |
| P-value correction for multiple comparisons | FDR method |
| Background list | Genes with >=10 reads per sample on average across all samples |
| Gene Scoring | DESeq2 Wald stat |
| Data availability | via DEE2 at accession SRP038101 (human) |
| Other parameters | Min gene set size of 5 |

```{r,mistake4a}

@@

```{r,mistake4b,fig.height=6,fig.width=6}

par(mar=c(2,20,2,0))

mres_sig_up_vals <- -log(mres_sig_up$pANOVA)
names(mres_sig_up_vals) <- mres_sig_up$set

mres_sig_dn_vals <- -log(mres_sig_dn$pANOVA)
names(mres_sig_dn_vals) <- mres_sig_dn$set

mres_es_up_vals <- mres_es_up$s.dist
names(mres_es_up_vals) <- mres_es_up$set

mres_es_dn_vals <- -1 * mres_es_dn$s.dist
names(mres_es_dn_vals) <- mres_es_dn$set

par(mfrow=c(2,1))

barplot( mres_sig_up_vals,
  horiz=TRUE,las=1,cex.names=0.65,col="gray",
  main="Up Sig (-log10 p-value)",
  xlab="-log10 p-value",
  cex.main=0.9)

barplot( mres_es_up_vals,
  horiz=TRUE,las=1,cex.names=0.65,col="gray",
  main="Up ES",
  xlab="ES",
  cex.main=0.9)

barplot( mres_sig_dn_vals,
  horiz=TRUE,las=1,cex.names=0.65,col="gray",
  main="Down Sig (-log10 p-value)",
  xlab="-log10 p-value",
  cex.main=0.9)

barplot( mres_es_dn_vals,
  horiz=TRUE,las=1,cex.names=0.65,col="gray",
  main="Down ES",
  xlab="-ES",
  cex.main=0.9)

par(mfrow=c(1,1))
par(mar=c(5.1,4.1,4.1,2.1))

## Mistake 5 - gene set list size too small or long

```{r,mistake5}

bg <- unique(sapply(strsplit(rownames(def)," "),"[[",2))

nrow(subset(def,padj<0.05 & log2FoldChange>0))
nrow(subset(def,padj<0.05 & log2FoldChange<0))
nrow(def)

sizes <- c(50,100,200,300,400,500,750,1000,1500,2000,2500,3000,4000,5000,6000,7000)

fup <- nrow(subset(fora(pathways=gsets, genes=up, universe=bg, minSize = 5, maxSize = Inf),padj<0.05))
fdn <- nrow(subset(fora(pathways=gsets, genes=dn, universe=bg, minSize = 5, maxSize = Inf),padj<0.05))

nups <- unlist(lapply(sizes, function(i) {
  genes <- unique(sapply(strsplit(rownames(head(subset(def, log2FoldChange>0),i))," "),"[[",2))
  fres <- fora(pathways=gsets, genes=genes, universe=bg, minSize = 5, maxSize = Inf)
  nrow(subset(fres,padj<0.05))
}))

ndns <- unlist(lapply(sizes, function(i) {
  genes <- unique(sapply(strsplit(rownames(head(subset(def, log2FoldChange<0),i))," "),"[[",2))
  fres <- fora(pathways=gsets, genes=genes, universe=bg, minSize = 5, maxSize = Inf)
  nrow(subset(fres,padj<0.05))
}))

df <- data.frame("up"=nups,"dn"=ndns,row.names=sizes)
df
plot(sizes,df$up,type="b",col="red",ylab="no. gene sets with FDR<0.05",xlab="gene list length")
lines(sizes,df$dn,type="b",col="blue")

plot(sizes,df$up,type="b",col="red",ylab="no. gene sets with FDR<0.05",xlab="gene list length",xlim=c(0,500))
lines(sizes,df$dn,type="b",col="blue")
